Staged Abstract Interpreters

### 2019-02-04 paper/10_intro.tex
reorganize and address typos

% \subsection{Motivation}

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to 
the input program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to realize 
Futamura projections, which first identifies the binding-time of variables in the program, i.e., 
they can be known whether statically or dynamically, and then evaluates the static part, and 
generates a residual program that relies on the dynamic part.
However, precisely analyzing binding-time for an arbitrary program is hard in general. As an 
alternative and pragmatic approach to specialization and partial evaluation, multi-stage 
programming (MSP for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage 
annotations (i.e., binding-time annotations) from the programmers. These staging annotations 
identify which part of the input program should be specialized.
As a classical example, we consider a power function: 
\begin{lstlisting}
def power(b: Rep[Int], e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If the program identifies that @b@ will be known dynamically (as shown in its type @Rep[Int]@ -- a representation of @Int@) 
and @e@ is statically known, say 5, 
then we can specialize the function \texttt{power} with respect to \texttt{e = 5} and generate a new 
specialized function that runs faster:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

Abstract interpretation \cite{DBLP:conf/popl/CousotC77} as a semantic approach to construct 
sound static analysis by approximation is almost as old as Futamura projections.
Some recent progress such as Abstracting Abstract Machines (AAM) uncovers a methodology to 
derive sound abstract interpreters from their concrete counterparts and has been applied to 
different variants of definitional interpreters and abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
Given the structural similarity between concrete and abstract interpreters, intellectually 
it is natural to raise the question whether it is possible to specialize an abstract interpreter, and how 
can we effectively do that having considered their different functionalities.
This paper studies the confluence of two old ideas --- Futamura project and abstract interpretation,
but from the perspective of their recent realizations --- multi-stage programming and definitional
abstract interpreters. To be specific, we present the application of the first Futamura projection
on abstract interpreters, and the approach of constructing optimizing abstract interpreters by 
multi-stage programming.

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of staging and abstraction.}
  \label{confluence}
\end{figure}

@@

On the other hand, the abstract interpreter, as a semantic artifact, should be written in a 
style that is easy for people to communicate the formulation and abstraction, but also can be 
implemented efficiently. As the slogan of multi-stage programming said, "abstraction without regret",
we draw connection between high-level description and efficient implementation of abstract interpreters,
just like the connection between concrete interpreters and compilers drawn by Futamura projection.
Particularly, we show a easy but systematic way of adding stage annotations to the abstract interpreter
without changing the code of interpreter skeleton, which is shared between four concrete or abstract,
unstaged or staged interpreters. We use LMS framework for staging, which allows only use types to 
annotate the binding-time. Therefore, the proposed approach bridges the gap between designing sound 
static analysis and implementing efficient program analyzer.

%Of course, all interpreters are metalinguistic abstractions, but some interpreters are more "abstract" than others \todo{maybe rephrase}. 
%Particularly, we also would like a systematic approach to optimize program analyzers and meanwhile minimally modifying the analyzer programs.
\iffalse
When staging a concrete interpreter, the programmers need to distinguish 
static and dynamic values --- the given program to be executed by the 
interpreter is classified as static because it is known at compile-time, 
and the inputs to that program are dynamic.  However, when staging an 
abstract interpreter, this distinction does not exist anymore. Because the 
abstract interpreter instantiates all the inputs as some form of abstract 
values, which are usually top elements in their abstract domains and are 
also statically known. 
Then what is the point of staging if there is no such distinction?
A surprising by-product of thinking about this question is to realize that 
we can apply the staged abstract interpreter on \textit{open} programs, and
the free variables representing other parts of the program (e.g., libraries) 
are dynamic inputs, therefore we obtain a modular analysis through staging, 
mechanically. 
\note{TR: I don't understand this. The program structure is static, the abstract values are still dynamic, no? They change in every iteration of the fixpoint algorithm} 
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor, but modular analysis has different asymptotics}
\fi

When targeting higher-order programs, either staged concrete interpreters or staged abstract interpreters
are able to compile a closure, i.e., specialize the call of interpreter with respect to the body
expression of the lambda term without knowing the actual argument value. By generalizing this observation,
we can actually specialize the abstract interpreter with any open programs, 
which unexpectedly leads to a modular analysis and improves scalability.
A open program contains free variables, which represent other parts of the program and will be left
as dynamic values.
For instances, one challenge in static analysis of modern software is that the programs are usually 
shipped with large library code \cite{toman_et_al:LIPIcs:2017:7121}, for example, it has been shown that 
analyzing a simple "Hello World" program in Java depends on additional 3,000 classes in the library 
\cite{DBLP:conf/oopsla/KulkarniMZN16}.
A precise whole-program analysis formulated as abstract interpreters can be very expensive
due to the scale of the program and the inherent complexity of the algorithm.
However, analyzing these libraries is sometimes unnecessarily repeated.
When applying the staged abstract interpreter on such open programs, we leave the unknown arguments and 
calling contexts as dynamic values and generate partial analyzing result which is 
represented as a residual program. The partial analyzing result can be reused and composed with the 
analyzing result of application code when available later. Therefore we mechanically obtain a modular
analysis through staging, even though the original algorithm is formulated as a whole-program analysis.

It has been observed that partially applying context-sensitivity on selected portion of the program could
improve the precision and efficiency \cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}.
We show that staging abstract interpreters as an approach to effectively implement hybird context-sensitivity\todo{}.

% Many optimization of static analysis are ad-hoc.

@@

\begin{itemize}
  \item Starting from a generic definition of interpreter for a higher-order functional language,
    whose syntax and semantics are described in Section~\ref{bg_lang}, we show its concrete
    interpreter (Section~\ref{unstaged_conc}), staged concrete interpreter (Section~\ref{stagedinterp}), 
    abstract interpreter (Section~\ref{unstaged_abs}) and staged abstract interpreter (Section~\ref{sai}).
    Our design allows one can derive these four different kinds of interpreters by mechanically changing
    the underlying semantics or stage annotations.
    %We later extend the staged abstract interpreter to support imperative language features (Section~\ref{cases_imp}), such as states and loops.
  \item We demonstrate that the staged abstract interpreter not only improves the \textit{efficiency}
    but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
    programs. \todo{Section X}.
  \item To demonstrate the applicability and usefulness of the proposed approach, we conduct three case studies 
    Section~\ref{cases_study}: 
    \subitem (1) We revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587} in Section~\ref{cs_ac}, 
    which was originally a closure generation technique applied on 0-CFA.
    We show that abstract compilation can be understood and easily implemented as an instance of our staged abstract interpreter.
    But unlike the abstract compilation, our approach does not require the implementers to rewrite the analysis.
    \subitem (2) We demonstrate pushdown control-flow analysis techniques (Section~\ref{cfa}), such as context-sensitive 
    analysis and more precise stores can be integrated to staged abstract interpreters without losing any precision.
    \subitem (3) We show that in an imperative language (Section~\ref{cases_imp}), by replacing the abstract domain
    to an interval domain (as well as staged), the staged abstract interpreter performs numerical analysis.
    % (Optional: taint/information flow analysis?)
  \item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
    for control-flow analysis (Section~\ref{evaluation}). 
    We compare our staged abstract interpreter with the unstaged version.
\end{itemize}

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections, which first identifies the binding-time of variables in
the program, i.e., they can be known whether statically or dynamically, and then
evaluates the static part, and generates a residual program that relies on the
dynamic part. However, precisely analyzing binding-time for an arbitrary program
is hard in general. As an alternative and pragmatic approach to specialization
and partial evaluation, multi-stage programming (MSP for short)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage annotations
(i.e., binding-time annotations) from the programmers. These staging annotations
identify which part of the input program should be specialized. As a classical
example, we consider a power function:

\begin{lstlisting}
def power(b: Rep[Int], e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}

If the program identifies that @b@ will be known dynamically (as shown in its
type @Rep[Int]@ -- a representation of @Int@) and @e@ is statically known, say
5, then we can specialize the function \texttt{power} with respect to \texttt{e
= 5} and generate a new specialized function that runs faster:

\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

Abstract interpretation \cite{DBLP:conf/popl/CousotC77} as a semantic approach
to construct sound static analysis by approximation is almost as old as Futamura
projections. Some recent progress such as Abstracting Abstract Machines (AAM)
uncovers a methodology to derive sound abstract interpreters from their concrete
counterparts and has been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}. Given the structural
similarity between concrete and abstract interpreters, intellectually it is
natural to raise the question whether it is possible to specialize an abstract
interpreter, and how can we effectively do that having considered their
different functionalities. This paper studies the confluence of two old ideas
--- Futamura projection and abstract interpretation, but from the perspective of
their recent realizations --- multi-stage programming and definitional abstract
interpreters. To be specific, we present the application of the first Futamura
projection on abstract interpreters, and the approach of constructing optimizing
abstract interpreters by multi-stage programming.

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of staging and approximation.}
  \label{confluence}
\end{figure}

@@

\todo{rewrite this paragraph}
On the other hand, the abstract interpreter, as a semantic artifact, should be
written in a style that is easy for people to communicate the formulation and
abstraction, but also can be implemented efficiently. As the slogan of
multi-stage programming said, "abstraction without regret", we draw connection
between high-level description and efficient implementation of abstract
interpreters, just like the connection between concrete interpreters and
compilers drawn by Futamura projection. Particularly, we show an easy but
systematic way of adding stage annotations to the abstract interpreter without
changing the code of the interpreter skeleton, which is shared between four concrete
or abstract, unstaged or staged interpreters. We use the LMS framework for staging,
which allows us only use types to annotate the binding-time. Therefore, the
proposed approach bridges the gap between designing sound static analysis and
implementing efficient program analyzer.

%Of course, all interpreters are metalinguistic abstractions, but some
%interpreters are more "abstract" than others \todo{maybe rephrase}.
%Particularly, we also would like a systematic approach to optimize program
%analyzers and meanwhile minimally modifying the analyzer programs.

\iffalse
When staging a concrete interpreter, the programmers need to distinguish static
and dynamic values --- the given program to be executed by the interpreter is
classified as static because it is known at compile-time, and the inputs to that
program are dynamic. However, when staging an abstract interpreter, this
distinction does not exist anymore. Because the abstract interpreter
instantiates all the inputs as some form of abstract values, which are usually
top elements in their abstract domains and are also statically known. Then what
is the point of staging if there is no such distinction? A surprising by-product
of thinking about this question is to realize that we can apply the staged
abstract interpreter on \textit{open} programs, and the free variables
representing other parts of the program (e.g., libraries) are dynamic inputs,
therefore we obtain a modular analysis through staging, mechanically.
\note{TR: I don't understand this. The program structure is static, the abstract values
are still dynamic, no? They change in every iteration of the fixpoint algorithm}
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor,
but modular analysis has different asymptotics}
\fi

When targeting higher-order programs, either staged concrete interpreters or
staged abstract interpreters are able to compile a closure, i.e., specialize the
call of interpreter with respect to the body expression of the lambda term
without knowing the actual argument value. By generalizing this observation, we
can actually specialize the abstract interpreter with any open programs, which
unexpectedly leads to a modular analysis and improves scalability. An open
program contains free variables, which represent other parts of the program and
will be left as dynamic values. For instance, one challenge in static analysis
of modern software is that the programs are usually shipped with large library
code \cite{toman_et_al:LIPIcs:2017:7121}, for example, it has been shown that
analyzing a simple "Hello World" program in Java depends on additional 3,000
classes in the library \cite{DBLP:conf/oopsla/KulkarniMZN16}. A precise
whole-program analysis formulated as abstract interpreters can be very expensive
due to the scale of the program and the inherent complexity of the algorithm.
However, analyzing these libraries is sometimes unnecessarily repeated. When
applying the staged abstract interpreter on such open programs, we leave the
unknown arguments and calling contexts as dynamic values and generate partial
analyzing result which is represented as a residual program. The partial
analyzing result can be reused and composed with the analyzing result of
application code when available later. Therefore we mechanically obtain a
modular analysis through staging, even though the original algorithm is
formulated as a whole-program analysis.

\iffalse
It has been observed that partially applying context-sensitivity on selected
portion of the program could improve the precision and efficiency
\cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}. We show that staging
abstract interpreters as an approach to effectively implement hybird
context-sensitivity\todo{}.
\fi

@@

\begin{itemize}
  \item Starting from a generic definition of an interpreter for a higher-order
    functional language, whose syntax and semantics are described in
    Section~\ref{bg_lang}, we show its concrete interpreter
    (Section~\ref{unstaged_conc}), staged concrete interpreter
    (Section~\ref{stagedinterp}), abstract interpreter (Section~\ref{unstaged_abs})
    and staged abstract interpreter (Section~\ref{sai}). Our design allows to
    derive these four different kinds of interpreters by mechanically changing the
    underlying semantics or stage annotations.
    %We later extend the staged abstract interpreter to support imperative
    %language features (Section~\ref{cases_imp}), such as states and loops.

  \item As shown in Section~\ref{sai}, we demonstrate that if we apply the
    abstract interpreters on open programs, it not only improves the
    \textit{efficiency} but also the \textit{scalability}, which are two major
    issues in static analysis.

  \item To demonstrate the applicability and usefulness of the proposed
    approach, we conduct three case studies in Section~\ref{cases_study}:

    \subitem (1) We revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587} in Section~\ref{cs_ac}, 
    which was originally a closure generation technique applied on 0-CFA.
    We show that abstract compilation can be understood and easily implemented
    as an instance of our staged abstract interpreter. But unlike abstract
    compilation, our approach does not require the implementers to rewrite the
    analysis.

    \subitem (2) We demonstrate pushdown control-flow analysis techniques
    (Section~\ref{cfa}), such as context-sensitive analysis and more precise
    stores can be integrated to staged abstract interpreters without losing any
    precision.

    \subitem (3) We show that in an imperative language
    (Section~\ref{cases_imp}), by using an interval domain as abstract domain
    (as well as staged), the staged abstract interpreter performs numerical 
    analysis.

    % (Optional: taint/information flow analysis?)
  \item We empirically evaluate the performance improvement by staging a
    big-step abstract interpreter for control-flow analysis
    (Section~\ref{evaluation}). We compare our staged abstract interpreter with the
    unstaged version.
\end{itemize}

### 2019-02-04 paper/21_bg_interp.tex
revise
reorganize and address typos

\subsection{Generic Interface} 
Before showing the actual implementation of the concrete semantics, let us first consider
a generic interface for the interpreters, which is semantics-agnostic and can be 
"interpreted" by either concrete or abstract, unstaged or staged semantics of our choice.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
It also declares initial values for environments and stores, and few abstract methods that 
manipulate environments and stores such as @put@ and @get@.

@@

The actual semantics and operations of the interpreter are left to be implemented later, 
these cinclude operations such as lifting literal terms to values (\texttt{close}, \texttt{num}), 
applying a closure (\texttt{apply\_closure}), branching (\texttt{branch0}),
and arithmetics (\texttt{prim\_eval}). We can choose to implement them in a way that is concrete or abstract,
unstaged or staged.
But, just based on these abstract operations, an expressive generic interpreter can be defined:

@@

\paragraph{Stage Polymorphism.}
Readers may notice that an abstract higher-kinded type \texttt{R[+T]} is defined in the trait, i.e. a
type-level function whici will be used to specify the binding-time, i.e., staged or not; the annotation 
\texttt{+} indicates that the type parameter \texttt{T} is covariant.
The LMS framework we use here allows us to specify binding-times purely based on types.
All occurrences of type \texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} 
essentially because they may potentially be dynamic when staging is introduced.
Obviously, the unstaged higher-kinded type then can be simply instantiated as an identity type function: 
\texttt{type R[+T] = T}. In Section~\ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{Finally tagless, poly embedding}

\subsection{Concrete Instantiation} \label{unstaged_conc}
Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, for the propose that properly models the heap, the type \texttt{Int} 
is used for address space \texttt{Addr}. The environment and store are just ordinary \texttt{Map}s in Scala.
A value can be either a tagged number \texttt{NumV} or a closure \texttt{CloV} that contains a lambda term
and an environment.

@@

we proceed to implement the components that define the semantics. 
As mentioned before, type \texttt{R} is concretized as an identity type function, thus
all types are just ordinary types.

>>>>

\subsection{Generic Interface}

Before showing the actual implementation of the concrete semantics, let us first
consider a generic interface for the interpreters, which is semantics-agnostic
and can be "interpreted" by either concrete or abstract, unstaged or staged
semantics of our choice. The trait \texttt{Semantics} first declares several
abstract type members, such as \texttt{Addr}, \texttt{Value}, \texttt{Env},
\texttt{Store} and returned type \texttt{Ans}. It also declares initial values
for environments and stores, and few abstract methods that manipulate
environments and stores such as @put@ and @get@.

@@

The actual semantics and operations of the interpreter are left to be
implemented later, these include operations such as lifting literal terms to
values (\texttt{close}, \texttt{num}), applying a closure
(\texttt{apply\_closure}), branching (\texttt{branch0}), and arithmetics
(\texttt{prim\_eval}). We can choose to implement them in a way that is concrete
or abstract, unstaged or staged. But, just based on these abstract operations,
an expressive generic interpreter can be defined:

@@

\paragraph{Stage Polymorphism}

Readers may notice that an abstract higher-kinded type \texttt{R[+T]} is defined
in the trait, i.e. a type-level function which will be used to specify the
binding-time, i.e., staged or not; the annotation \texttt{+} indicates that the
type parameter \texttt{T} is covariant. The LMS framework we use here allows us
to specify binding-times purely based on types. All occurrences of type
\texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} essentially
because they may potentially be dynamic when staging is introduced. Obviously,
the unstaged higher-kinded type then can be simply instantiated as an identity
type function: \texttt{type R[+T] = T}. In Section~\ref{stagedinterp},
\texttt{R} will be instantiated as a staging annotation.

\todo{Finally tagless, poly embedding}

\subsection{Concrete Instantiation} \label{unstaged_conc}

Now we can instantiate a concrete interpreter. We first concretize the abstract
types as follows. Since the interpreter is concrete, for the purpose that
properly models the heap, the type \texttt{Int} is used for address space
\texttt{Addr}. The environment and store are just ordinary \texttt{Map}s in
Scala. A value can be either a tagged number \texttt{NumV} or a closure
\texttt{CloV} that contains a lambda term and an environment.

@@

We proceed to implement the components that define the semantics. As mentioned
before, type \texttt{R} is concretized as an identity type function, thus all
types are just ordinary types.

### 2019-02-04 paper/23_bg_stage.tex
reorganize and address typos

\subsection{Multi-Stage Programming in LMS}
Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a multi-stage programming framework implemented 
as a Scala library that enables dynamic code generation in a type-safe manner.
Different from the syntactic approach in MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} that uses quotations, 
LMS distinguishes binding-time solely based on types.
LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary type, indicating an expression will be 
known at next stage. All operations acting on a @Rep[T]@ expression will be residualized in the generated code. However,
we need to provide intermediate representation and code generation support for type @T@ to LMS. 
Fortunately, the LMS framework already provides
such support for primitive types and commonly used data structure such as arrays and maps; implementing such support
for custom class is also straightforward. Let's go back and see how LMS can be used to specialize the power function
we mentioned in Section 1.

@@

In the code shown above, \texttt{power} takes two arguments where \texttt{b} is declared as \texttt{Rep[Int]} type 
meaning that \texttt{b} is a representation of \texttt{Int} but whose value will be available at the next stage. 
Meanwhile, the result of the @power@ function will be available at the next stage as well, thus the return value 
of \texttt{power} is also of type \texttt{Rep[Int]}. 
Then we specialize @power@ in function @snippet@ by providing @x@ to @5@.
The generated code for \texttt{power5} is a function that takes only one argument which corresponds to @b@,
and the body of this function multiples @b@ five times.

\subsection{Staged Concrete Semantics} 
In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations
in trait @RepConcInterpOps@.
Meanwhile, it is also extended from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures.
To stage the interpreter, we first identifies the syntactic input is known statically, 
and the interpreter returns a pair of values and store, which both becomes dynamic values.
Thus the staged stores propogate the binding-time to environments, because they share the same
address component.

@@

Next, @apply_closure@ takes three arguments: the function value, argument value and the latest store, 
which are both of @Rep@ type. This means basically there is nothing we can do at the current stage --- 
so we create a new node @ApplyClosure@ that contains these arguments in the intermediate representation graph,
and in the code generation phase we can emit the code that performs the application. The additional @reflectEffect@
is a function in LMS that handles side effects.

@@

\paragraph{Code Generation.}
When generating code for the next stage, values like @CompiledClo@ and @NumV@ are kept as they are because
they are intended exist in the next stage; however, we need to treat @ApplyClosure@ a little differently.

@@

Concretely, as shown in the code above, when emitting code for @ApplyClosure@, we just 
emit a function application where the function value is extracted from the @CompiledClo@ object
(we assume the input program is well-typed so that we can safely cast it to @CompiledClo@).
Argument and store are the remaining rest fields in @ApplyClosure@.

>>>>

\subsection{Multi-Stage Programming in LMS}

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the syntactic
approach in MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} that uses quotations,
LMS distinguishes binding-time solely based on types. LMS provides a type
constructor @Rep[T]@ where @T@ can be an arbitrary type, indicating an
expression will be known at next stage. All operations acting on a @Rep[T]@
expression will be residualized in the generated code. However, we need to
provide intermediate representation and code generation support for type @T@ to
LMS. Fortunately, the LMS framework already provides such support for primitive
types and commonly used data structure such as arrays and maps; implementing
such support for custom class is also straightforward. Let's go back and see how
LMS can be used to specialize the power function we mentioned in Section 1.

@@

In the code shown above, \texttt{power} takes two arguments where \texttt{b} is
declared as \texttt{Rep[Int]} type meaning that \texttt{b} is a representation
of \texttt{Int} but whose value will be available at the next stage. Meanwhile,
the result of the @power@ function will be available at the next stage as well,
thus the return value of \texttt{power} is also of type \texttt{Rep[Int]}. Then
we specialize @power@ in function @snippet@ by providing @x@ to @5@. The
generated code for \texttt{power5} is a function that takes only one argument
which corresponds to @b@, and the body of this function multiplies @b@ five
times.

\subsection{Staged Concrete Semantics}

In the staging part, we share the same concrete type instantiation, i.e., trait
\texttt{Concrete}, but we reimplement the staged version for the concrete
operations in trait @RepConcInterpOps@. Meanwhile, it is also extended from
@LMSOps@ which provides necessary support for the staged version of primitive
types and data structures. To stage the interpreter, we first identify the
syntactic input that is known statically, and the interpreter returns a pair of
values and store, which both becomes dynamic values. Thus the staged stores
propagate the binding-time to environments, because they share the same address
component.

@@

Next, @apply_closure@ takes three arguments: the function value, argument value
and the latest store, which are all of @Rep@ type. This means basically there
is nothing we can do at the current stage --- so we create a new node
@ApplyClosure@ that contains these arguments in the intermediate representation
graph, and in the code generation phase we can emit the code that performs the
application. The additional @reflectEffect@ is a function in LMS that handles
side effects.

@@

\paragraph{Code Generation}

When generating code for the next stage, values like @CompiledClo@ and @NumV@
are kept as they are because they are intended exist in the next stage; however,
we need to treat @ApplyClosure@ a little differently.

@@

Concretely, as shown in the code above, when emitting code for @ApplyClosure@,
we just emit a function application where the function value is extracted from
the @CompiledClo@ object (we assume the input program is well-typed so that we
can safely cast it to @CompiledClo@). Argument and a store are the remaining rest
fields in @ApplyClosure@.

### 2019-02-04 paper/30_sai.tex
reorganize and address typos

\subsection{Staged Lattices} 
In Section~\ref{stagedpoly_lat}, we exploited the higher-kinded type @R@
to leave space for staging lattices, now we instantiate the type @R@ to @Rep@ and 
still use powersets as an example to present its staged version.

@@

The type parameter @T:Typ@ of @RepLattice@ requires that the elements of sets
is can also be staged. Otherwise, without knowing how to stage the elements in the set,
we can not stage the set either.
The methods defined operate on type @Rep[Set[T]]@, thus the underlying implementation
such as @union@ and @intersect@ will be mapped to a node in the IR graph during staging and 
eventually emitted in the generated code. 
Again, other structures such as maps and tuples are implemented in a similar way.

\subsection{Staged Abstract Semantics} 
We have seen how to obtain a staged concrete semantics based on types, 
now we take the same approach to obtain a staged abstract semantics.
The basic operations are largely kept the same as in the unstaged version, except the types are changed to @Rep@.
Besides, when we update the environment, the identifier @x@ is known statically, but the environment map
has type @Rep[Map[Ident,Addr]]@, so we apply @lift@ to @x@ to turn it as a next-stage value.

@@

Now we have already obtained an end-to-end staged abstract interpreter that is able to 
specialize an analysis. However, we treat the data structures such as @Map@s as black-boxes,
which means any operations on a @Map@ becomes code in the next stage.
But, as we identified when introducing the generic interface, the keys of any environment maps are 
identifiers in the program, which are completely known statically. 
This leaves us a chance to further specialize the data structures. 
Assume the @Map[K,V]@ is implemented as a hash map, if the keys $K$ are known, then the indices can be computed
statically. Thus the specialized map would be an array @Array[Rep[V]]@ whose elements 
are next-stage values; all the accesses to the array is determined during staging.

@@

\subsection{Discussion}
%We have presented all of our four different interpreters as shown in Figure~\ref{confluence}.
In the literature of partial evaluation, \citeauthor{10.1007/3-540-61580-6_11} provided guidelines
on what to do and not to do when specializing a concrete interpreter \cite{10.1007/3-540-61580-6_11}. 
We borrow these guidelines and extend them to abstract interpreters. We discuss decisions we made to 
achieve this and examine some alternatives.

\paragraph{Big-step vs Small-step.}
What we implemented is a big-step, compositional abstract interpreter, 
where "compositional" means that every recursive call of our abstract interpreter is applied to proper 
substructures of the current syntactic parameters \cite{10.1007/3-540-61580-6_11}.
This compositionality ensures that specialization can be done by unfolding,
as well as that the specialization procedure terminates.
Nevertheless, it is also possible to specialize a small-step operational abstract semantics ---
\citeauthor{Johnson:2013:OAA:2500365.2500604}
showed this in abstract compilation \cite{Boucher:1996:ACN:647473.727587} style as one of their optimizations
of Abstract Abstract Machines \cite{Johnson:2013:OAA:2500365.2500604}. 
However, the generated abstract bytecode still requires another small-step abstract machine 
to execute, which is an additional engineering efforts.
Another alternative approach for efficient specialization is to write the abstract interpreter 
in a big-step, monadic style \cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

\paragraph{Correctness and Soundness.}
Based on the assumption that LMS preserves the semantics during staging,
we are confident that the staged abstract interpreter does the same analysis compared 
with the unstaged one. Moreover, the optimization done by staging does not compromise
any soundness.

>>>>

\subsection{Staged Lattices}

In Section~\ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
leave space for staging lattices, now we instantiate the type @R@ to @Rep@ and
still use powersets as an example to present its staged version.

@@

The type parameter @T:Typ@ of @RepLattice@ requires that the elements of sets
can also be staged. Otherwise, without knowing how to stage the elements in the
set, we can not stage the set either. The methods defined operate on type
@Rep[Set[T]]@, thus the underlying implementation such as @union@ and
@intersect@ will be mapped to a node in the IR graph during staging and
eventually emitted in the generated code. Again, other structures such as maps
and tuples are implemented in a similar way.

\subsection{Staged Abstract Semantics}

We have seen how to obtain a staged concrete semantics based on types, now we
take the same approach to obtain a staged abstract semantics. The basic
operations are largely kept the same as in the unstaged version, except the
types are changed to @Rep@. Besides, when we update the environment, the
identifier @x@ is known statically, but the environment map has type
@Rep[Map[Ident,Addr]]@, so we apply @lift@ to @x@ to turn it as a next-stage
value.

@@

Now we have already obtained an end-to-end staged abstract interpreter that is
able to specialize an analysis. However, we treat the data structures such as
@Map@s as black-boxes, which means any operations on a @Map@ become code in the
next stage. But, as we identified when introducing the generic interface, the
keys of any environment maps are identifiers in the program, which are
completely known statically. This leaves us a chance to further specialize the
data structures. Assume the @Map[K,V]@ is implemented as a hash map, if the keys
$K$ are known, then the indices can be computed statically. Thus the specialized
map would be an array @Array[Rep[V]]@ whose elements are next-stage values; all
the accesses to the array is determined during staging.

@@

\subsection{Discussion}

In the literature of partial evaluation, \citeauthor{10.1007/3-540-61580-6_11}
provided guidelines on what to do and not to do when specializing a concrete
interpreter \cite{10.1007/3-540-61580-6_11}. We borrow these guidelines and
extend them to abstract interpreters. We discuss decisions we made to achieve
this and examine some alternatives.

\paragraph{Big-step vs Small-step}

What we implemented is a big-step, compositional abstract interpreter, where
"compositional" means that every recursive call of our abstract interpreter is
applied to proper substructures of the current syntactic parameters
\cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as that the specialization
procedure terminates. Nevertheless, it is also possible to specialize a
small-step operational abstract semantics ---
\citeauthor{Johnson:2013:OAA:2500365.2500604} showed this in abstract
compilation \cite{Boucher:1996:ACN:647473.727587} style as one of their
optimizations of Abstract Abstract Machines
\cite{Johnson:2013:OAA:2500365.2500604}. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering efforts. Another alternative approach for efficient
specialization is to write the abstract interpreter in a big-step, monadic style
\cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

\paragraph{Correctness and Soundness}

Based on the assumption that LMS preserves the semantics during staging, we are
confident that the staged abstract interpreter does the same analysis compared
with the unstaged one. Moreover, the optimization done by staging does not
compromise any soundness.
